They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-15 paper/paper.Rmd
i -> r, new table

The general formula for the probability $P_E(i)$ that a best-of-$s$ series lasts $i$ rounds given constant probability $p$ of one team[^11] winning each round is

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq-series-length}{}{
P_E(i) = \frac{(i - 1)!}{s_1!(i - s_2)!}(p^{s_2}q^{s_2^{'}} + p^{s_2^{'}}q^{s_2}).
}\label{eq:series-length}\end{equation}
```
```{r}
#| label: cod_series_outcome_prop
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_n_games <- function(n, p = 0.5, max_round = .max_round) {
  s <- ceiling(max_round / 2)
  (factorial(n - 1) / (factorial(s - 1) * factorial(n - s))) * (p^s * (1 - p)^(n - s) + p^(n - s) * (1 - p)^s)
}

@@

```{r}
#| label: cod_series_outcome_prop-table
cod_streaks_naive_res$series |>
  select(
    n_rounds,
    prop_expected,
    prop_actual,
    n_actual
  ) |>
  arrange(n_rounds) |>
  gt::gt() |>
  gt::cols_label(
    n_rounds = 'Series lasts\ni rounds',
    prop_expected = 'P_{E}(i)',
    prop_actual = 'P_{O}(i)',
    n_actual = 'N_O'
  ) |>
  gt::fmt_percent(
    decimals = 1,
    columns = c('prop_expected', 'prop_actual')
  ) |>
  gt::tab_spanner(
    label = 'Observed',
    columns = c('prop_actual', 'n_actual')
  ) |>
  gt::as_latex()
```

Table \ref{tbl:cod-prob-series-lasting-i-rounds} shows the expected probabilities for $s = 11$ under the naive assumption that $p_0 = 0.5$ for a given round, along with the observed proportions $P_O(i)$ in CoD SnD.

```{=tex}
\begin{longtable}{rrrr}
\caption{The probabilities that a best-of-11 series lasts $i$ rounds ($P_E(i)$, where $i \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(i)$ and as a proportion $P_O(i)$ of all series.}\label{tbl:cod-prob-series-lasting-i-rounds} \\
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}
```
```{r}
#| label: cod_series_outcomes_naive_chi
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11].
}\label{eq:chi-squ-test}\end{equation}
```
as 16.0 (p-value of 0.0068), I can comfortably reject the constant probability null hypothesis for $p_0 = 0.5$, event at a confidence level of $\alpha = 0.01$.

@@

cod_streaks_other_res |>
  filter(idx %in% c(0L, 2L)) |> 
  mutate(
    lab = sprintf('$p$ = %s', p)
  ) |>
  select(lab, series_streaks) |>
  unnest(series_streaks) |>
  select(lab, n_rounds, prop_expected) |>
  mutate(
    across(prop_expected, ~scales::percent(.x, accuracy = 0.1))
  ) |>
  pivot_wider(
    names_from = lab,
    values_from = prop_expected
  ) |>
  inner_join(
    cod_streaks_naive_res$series |>
      transmute(n_rounds, `$P_O(i)$` = scales::percent(prop_actual, accuracy = 0.1)),
    by = 'n_rounds'
  ) |>
  arrange(n_rounds) |>
  rename(`Series lasts $i$ rounds` = n_rounds) |>
  knitr::kable()
```

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & \(p\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(i)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $p_2 = 0.575$.[^12] I observe that $P_E(i)$ is notably larger for $p_{2}$ when $i \in [6, 7]$, more closely matching $P_O(i)$. $p_{2}$ is also closer to $P_O(i)$ for $i \in [9, 10]$, although not for $i \in [8, 11]$.

@@

```{=tex}
\begin{longtable}[]{@{}rrrr@{}}
\caption{The probabilities that a best-of-11 series lasts $i$ rounds under the naive assumption $p_0 = 0.5$ and the maximum likelihood estimation $p_2 = 0.575$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
Series lasts \(i\) rounds & \(p_0\) = 0.5 & \(p_2\) = 0.575 & \(P_O(i)\) \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}
```
It is fair to conclude that the constant round win probability assumption can be valid in CoD SnD series with the appropriate choice of $p$ ($\approx 0.575$).

@@

```{r}
#| label: cod_round_win_prop_after_k_wins-table
cod_round_win_prop_after_k_wins |> 
  mutate(
    across(
      matches('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  # filter(win_streak == 3) |> 
  select(
    win_streak,
    win_series,
    n_rounds,
    total,
    prop,
    prior_p,
    expected_poststreak_p,
    gt_p_value,
    gt_p_value_adj
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    decimals = 0,
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::cols_label(
    win_streak = 'Streak',
    win_series = 'Win Series?',
    n_rounds = 'Series lasts $i$ rounds',
    total = 'N',
    prop = 'Obs.',
    prior_p = 'Naive',
    expected_poststreak_p = 'Discounted',
    gt_p_value = 'Naive',
    gt_p_value_adj = 'Discounted'
  ) |> 
  gt::tab_spanner(
    label = '%s',
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::tab_spanner(
    label = 'Binom. > p-value',
    columns = c(gt_p_value, gt_p_value_adj)
  )
```

Despite the implication that one can model team round win probability in CoD SnD as a constant, anecdotally many believe in momentum i.e. streaks not due to random chance. I now evaluate win frequency in rounds following a streak of length $k$. I account for the length of the series ($i$ rounds), modeling the expected round win probability after a streak of length $k$ as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pki}{}{
  \hat{P}_{wki} = \hat{P}(\text{win round | k-round winning streak, series lasting i rounds})=\{
  \begin{array}{cl}
    \frac{6 - k}{i - k}, & \text{team wins series} \\
    \frac{11 - 6 - k}{i - k}, & \text{team loses series}
  \end{array}.
}\label{eq:pki}\end{equation}
```
$\hat{P}_{lki} = \hat{P}(\text{lose round | k-round losing streak, series lasting i rounds})$ can be formulated symmetrically.

Miller et al. [-@miller2018] provide evidence for negative recency suggesting that

```{=tex}
\begin{equation}\protect\hypertarget{eq:pkid}{}{
  \begin{array}{rcl}
  \hat{P}_{ki,\text{diff}} & = & \hat{P}(\text{win round | k-round winning streak, series lasting i rounds}) -  \\
    & & \hat{P}(\text{lose round | k-round losing streak, series lasting i rounds}) \\
    E[\hat{P}_{ki,\text{diff}}] & \neq & 0
  \end{array}
}\label{eq:pkid}\end{equation}
```
given a deterministic i.i.d. Bernoulli probability (Equation \ref{eq:pki}.[^13]

@@

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on round win percentages in rounds immediately following streaks of three wins.[^14] Table \ref{tbl:cod-round-win-prop-after-k-wins} shows $\hat{P}_{wki}$ for $k = 3$ winning streaks, for both teams that empirically won and lost the series. (Some combinations of $i$ and $\text{team loses series}$ are irrelevant, such as the case when $i \in [7, 8, 9]$, so these combinations are excluded.)

@@

# References {.unnumbered}

>>>>

The general formula for the probability $P_E(r)$ that a best-of-$s$ series lasts $r$ rounds given constant probability $p$ of one team[^11] winning each round is

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq-series-length}{}{
P_E(r) = \frac{(r - 1)!}{s_1!(r - s_2)!}(p^{s_2}q^{s_2^{'}} + p^{s_2^{'}}q^{s_2}).
}\label{eq:series-length}\end{equation}
```
```{r}
#| label: cod_series_outcome_prop
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_n_games <- function(n, p = 0.5, max_round = .max_round) {
  s <- ceiling(max_round / 2)
  (factorial(n - 1) / (factorial(s - 1) * factorial(n - s))) * (p^s * (1 - p)^(n - s) + p^(n - s) * (1 - p)^s)
}

@@

```{r}
#| label: cod_series_outcome_prop-table
cod_streaks_naive_res$series |>
  select(
    n_rounds,
    prop_expected,
    prop_actual,
    n_actual
  ) |>
  arrange(n_rounds) |>
  gt::gt() |>
  gt::cols_label(
    n_rounds = 'Series lasts\nr rounds',
    prop_expected = 'P_{E}(r)',
    prop_actual = 'P_{O}(r)',
    n_actual = 'N_O'
  ) |>
  gt::fmt_percent(
    decimals = 1,
    columns = c('prop_expected', 'prop_actual')
  ) |>
  gt::tab_spanner(
    label = 'Observed',
    columns = c('prop_actual', 'n_actual')
  ) |>
  gt::as_latex()
```

Table \ref{tbl:cod-prob-series-lasting-r-rounds} shows the expected probabilities for $s = 11$ under the naive assumption that $p_0 = 0.5$ for a given round, along with the observed proportions $P_O(r)$ in CoD SnD.

```{=tex}
\begin{longtable}{rrrr}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds ($P_E(r)$, where $r \in [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50\% probability ($p_0 = 0.5$) of winning each game. The observed frequencies observed are shown as a count $N_O(r)$ and as a proportion $P_O(r)$ of all series.}\label{tbl:cod-prob-series-lasting-r-rounds} \\
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
Series lasts $r$ rounds & $P_E(r)$ & $P_O(r)$ & $N_O(r)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}
```
```{r}
#| label: cod_series_outcomes_naive_chi
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq:series-length}{}{
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(r) - P_E(r))^2}{P_E(r)}, r \in R = [6, 7, 8, 9, 10, 11].
}\label{eq:chi-squ-test}\end{equation}
```
as 16.0 (p-value of 0.0068), I can comfortably reject the constant probability null hypothesis for $p_0 = 0.5$, event at a confidence level of $\alpha = 0.01$.

@@

cod_streaks_other_res |>
  filter(idx %in% c(0L, 2L)) |> 
  mutate(
    lab = sprintf('$p$ = %s', p)
  ) |>
  select(lab, series_streaks) |>
  unnest(series_streaks) |>
  select(lab, n_rounds, prop_expected) |>
  mutate(
    across(prop_expected, ~scales::percent(.x, accuracy = 0.1))
  ) |>
  pivot_wider(
    names_from = lab,
    values_from = prop_expected
  ) |>
  inner_join(
    cod_streaks_naive_res$series |>
      transmute(n_rounds, `$P_O(r)$` = scales::percent(prop_actual, accuracy = 0.1)),
    by = 'n_rounds'
  ) |>
  arrange(n_rounds) |>
  rename(`Series lasts $r$ rounds` = n_rounds) |>
  knitr::kable()
```

@@

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
\caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endfirsthead
\toprule()
Method & \(p(r)\) & \(\chi^2\) (p-value) \\
\midrule()
\endhead
0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
1. Method of moments & 0.5725 & 3.6 (0.6) \\
2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
\bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(r)$ when re-applying Equation \ref{eq:series-length} for the maximum likelihood estimate $p_2 = 0.575$.[^12] I observe that $P_E(r)$ is notably larger for $p_{2}$ when $r \in [6, 7]$, more closely matching $P_O(r)$. $p_{2}$ is also closer to $P_O(r)$ for $r \in [9, 10]$, although not for $r \in [8, 11]$.

@@

```{=tex}
\begin{longtable}[]{@{}rrrr@{}}
\caption{The probabilities that a best-of-11 series lasts $r$ rounds under the naive assumption $p_0 = 0.5$ and the maximum likelihood estimation $p_2 = 0.575$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
Series lasts \(r\) rounds & \(p_0(r)\) = 0.5 & \(p_2(r)\) = 0.575 & \(P_O(r)\) \\
\midrule()
\endhead
6 & 3.1\% & 4.2\% & 4.7\% \\
7 & 9.4\% & 11.2\% & 11.9\% \\
8 & 16.4\% & 17.8\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.0\% & 21.5\% \\
11 & 24.6\% & 22.0\% & 23.7\% \\
\bottomrule()
\end{longtable}
```
It is fair to conclude that the constant round win probability assumption can be valid in CoD SnD series with the appropriate choice of $p$ ($\approx 0.575$).

@@

```{r}
#| label: cod_round_win_prop_after_k_wins-table
cod_round_win_prop_after_k_wins |> 
  mutate(
    across(
      matches('p_value'), 
      ~case_when(
        .x <= 0.01 ~ '<=0.01', 
        round(.x, 2) == 1 ~ '1',
        TRUE ~ as.character(round(.x, 2))
      )
    )
  ) |>
  filter(win_streak == 3) |> 
  arrange(win_streak, n_rounds, win_series) |> 
  transmute(
    # win_streak,
    n_rounds,
    across(win_series, ~ifelse(.x, 'win', 'lose')),
    total,
    prop,
    prior_p,
    expected_poststreak_p,
    gt_p_value,
    gt_p_value_adj
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    decimals = 0,
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::cols_label(
    # win_streak = 'Streak',
    win_series = 'Win Series?',
    n_rounds = 'Series lasts r rounds',
    total = 'N',
    prop = 'O',
    prior_p = 'a',
    expected_poststreak_p = 'b',
    gt_p_value = 'a',
    gt_p_value_adj = 'b'
  ) |> 
  gt::tab_spanner(
    label = 'x',
    columns = c(prop, prior_p, expected_poststreak_p)
  ) |> 
  gt::tab_spanner(
    label = 'y',
    columns = c(gt_p_value, gt_p_value_adj)
  ) |> 
  gt::tab_footnote(
    footnote = 'Binom',
    locations = gt::cells_column_labels(columns = c(gt_p_value))
  ) |> 
  gt::as_latex()
```

Despite the implication that one can model team round win probability in CoD SnD as a constant, anecdotally many believe in momentum i.e. streaks not due to random chance. I now evaluate win frequency in rounds following a streak of length $k$. I account for the length of the series ($r$ rounds), modeling the expected round win probability after a streak of length $k$ as

```{=tex}
\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  \hat{p}^{kr,\text{win}}_E = \hat{p}(\text{win round | k-round winning streak, series lasting r rounds})=\{
    \begin{array}{cl}
    \frac{6 - k}{i - k}, & \text{team wins series} \\
    \frac{11 - 6 - k}{i - k}, & \text{team loses series}
    \end{array}.
  }\label{eq:pwkr}
\end{equation}
```
$\hat{p}^{kr,loss}_E = \hat{p}(\text{lose round | k-round losing streak, series lasting r rounds})$ can be formulated symmetrically.

While one might think that $E[\hat{p}^{kr,\text{win}} - \hat{p}^{kr,\text{loss}}] = 0$, Miller et al. [-@miller2018] suggest that, given a deterministic i.i.d. Bernoulli probability (Equation \ref{pwkr}, this is not true due to streak-selection bias.[^13][^14]

@@

@@

I don't attempt to measure this expectation like Miller et al. [-@miller2018], who perform their analysis longitudinally per study participant; nonetheless, I apply their formulation for applying the "downward bias" on post-streak success to come up with estimates $\hat{P}_{kr,\text{win}}^{\text{adj.}}$, specifying $r$ as the number of trials.

Given that people typically perceive streaks as beginning after the third success (or failure) at minimum [@carlson2007], I focus on round win percentages in rounds immediately following streaks of three wins.[^15] Table \ref{tbl:cod-round-win-prop-after-k-wins} shows $\hat{P}_{kr,\text{win}}$ for $k = 3$ winning streaks, for both teams that empirically won and lost the series. (Some combinations of $r$ and $\text{team loses series}$ are irrelevant, such as the case when $r \in [7, 8, 9]$, so these combinations are excluded.)

[^15]: Three happens to also be a reasonable number for series that lasts at maximum 11 rounds.

```{=tex}
\begin{longtable}{rrrrrrrr}
\toprule
&  &  & \multicolumn{3}{c}{$\hat{p}^{kr,\text{win}}$} & \multicolumn{2}{c}{$p$-value} \\
\cmidrule(lr){4-6} \cmidrule(lr){7-8}
Series lasts $r$ rounds & Win Series? & $N^{kr,\text{win}}_O$ & $O$ & $E$ & $\text{adj.}$ & $E$ & $\text{adj.}$ \\ 
\midrule
7 & win & 209 & $75\%$ & $75\%$ & $61\%$ & 0.58 & <=0.01 \\ 
8 & win & 209 & $62\%$ & $60\%$ & $45\%$ & 0.28 & <=0.01 \\ 
9 & win & 193 & $52\%$ & $50\%$ & $35\%$ & 0.33 & <=0.01 \\ 
10 & lose & 60 & $13\%$ & $14\%$ & $10\%$ & 0.64 & 0.26 \\ 
10 & win & 151 & $44\%$ & $43\%$ & $29\%$ & 0.45 & <=0.01 \\ 
11 & lose & 129 & $24\%$ & $25\%$ & $16\%$ & 0.63 & <=0.01 \\ 
11 & win & 150 & $40\%$ & $38\%$ & $25\%$ & 0.29 & <=0.01 \\ 
\bottomrule
\end{longtable}
```
# References {.unnumbered}
